---
title: "Cyclistic Bike Share - A Case Study"
author: "Aurin Bose"
date: "2023-01-06"
output:
  html_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

## I Introduction

The following Case Study has been done in order to determine how Annual
member differ from casual riders on a fictional bike sharing service
called Cyclistic which in 2016 launched a successful bike-share
offering. Since then, the program has grown to a fleet of 5,824 bicycles
that are geotracked and locked into a network of 692 stations across
Chicago. The bikes can be unlocked from one station and returned to any
other station in the system anytime.

The analysis performed on their bike sharing data will help them
determine how to convert casual riders into annual members which in turn
will be the key to their future growth. The six key phases of data
analysis have been used to complete this objective and will be
highlighted in the rest of this report.

## II Ask Phase

The Ask Phase is used by data analysts to determine the problem to be
solved and how their insights could drive business decisions.

In this case the central query to be answered has been assigned to us
namely: How do annual members and casual riders use Cyclistic bikes
differently?

## III Prepare Phase

The Prepare Phase is meant for collecting, organizing, sorting and
filtering data. It is important in this phase to check where the data
comes from, whether it meets the ROCCC principles and if the data can be
used for the purpose of data analysis.

The data used here is based on the Divvy case study "'Sophisticated,
Clear, and Polished': Divvy and Data Visualization" written by Kevin
Hartman found
[here](https://artscience.blog/home/divvy-dataviz-case-study).

(*Note*: The datasets have a different name because Cyclistic is a
fictional company. For the purposes of this case study, the datasets are
appropriate and will enable you to answer the business questions. The
data has been made available by Motivate International Inc. under this
[license](https://www.divvybikes.com/data-license-agreement))

Since the data has been obtained from a first hand source, it can
classified as trustworthy and any licensing issues have been addressed
as well. The data provided contains bike ride information from the past
12 months from when this Case Study is being performed, hence January to
December 2022.

## IV Process Phase

The Process Phase is used to check the data for errors, select the tools
we wish to use for analysis, transform the data as per our requirements
and clean the data.

For the purposes of this study since the dataset provided was of
significant size, the best options were to use SQL or R programming and
subsequently R has been deemed most appropriate.

The first step in R is to load the libraries necessary in the entire
analysis process:

```{r}
library(tidyverse)
library(lubridate) 
library(ggplot2)
library(tidyr)
library(scales)
```

Then we check the current directory in R and set it to the folder where
our dataset is located if it is not assigned already.

Next we can begin loading all our dataset into R which has been stored
in the form of CSV files:

```{r}
jan_22<- read.csv("202201-divvy-tripdata.csv")
feb_22<- read.csv("202202-divvy-tripdata.csv")
mar_22<- read.csv("202203-divvy-tripdata.csv")
apr_22<- read.csv("202204-divvy-tripdata.csv")
may_22<- read.csv("202205-divvy-tripdata.csv")
jun_22<- read.csv("202206-divvy-tripdata.csv")
jul_22<- read.csv("202207-divvy-tripdata.csv")
aug_22<- read.csv("202208-divvy-tripdata.csv")
sep_22<- read.csv("202209-divvy-tripdata.csv")
oct_22<- read.csv("202210-divvy-tripdata.csv")
nov_22<- read.csv("202211-divvy-tripdata.csv")
dec_22<- read.csv("202212-divvy-tripdata.csv")

```

Once loaded we can preview the columns in our data and check their basic
structures as shown in the samples below:

```{r}
colnames(jan_22)
colnames(feb_22)

```

```{r}
str(jan_22)
str(feb_22)
```

Now we can merge our individual datasets into a large dataset that will
be used for the overall analysis process:

```{r}
all_trip<- bind_rows(jan_22, feb_22, mar_22, apr_22, may_22, jun_22, jul_22, 
                     aug_22, sep_22, oct_22, nov_22, dec_22)
```

Then we can preview the new merged data set by performing the below
operations:

```{r}
head(all_trip) #see the first 6 rows of data frame
colnames(all_trip) #list of column names 
str(all_trip) #see list of columns and data types
summary(all_trip) #statistical summary of data

all_trip<- all_trip %>% rename (member_type=member_casual)
```

Before we begin our analysis there are a few of steps to be completed so
that our data clean and suitable for making calculations:

-   Firstly the formats of the started_at and ended_at columns were
    changed to Year-month-day and Hour-minute-seconds.

```{r}
all_trip$started_at <- as.POSIXct(
  all_trip$started_at, 
  format = "%Y-%m-%d %H:%M:%S"
)

all_trip$ended_at <- as.POSIXct(
  all_trip$ended_at, 
  format = "%Y-%m-%d %H:%M:%S"
)

all_trip<- all_trip %>% arrange(started_at)
```

-   Then a new column was created called ride length from calculating
    the difference between ride start time and end time.

```{r}
all_trip$ride_length <- difftime(all_trip$ended_at,all_trip$started_at)
all_trip$ride_length <- as.numeric(as.character(all_trip$ride_length))
is.numeric(all_trip$ride_length)
```

-   Next separate columns were created for date, year, month, day, day
    of the week and even time of day.

```{r}
all_trip$date <- as.Date(all_trip$started_at)
all_trip$month <- format(as.Date(all_trip$date), "%m")
all_trip$day <- format(as.Date(all_trip$date), "%d")
all_trip$year <- format(as.Date(all_trip$date), "%Y")
all_trip$day_of_week <- format(as.Date(all_trip$date), "%A")
all_trip$YMD <- format(all_trip$started_at, "%Y-%m-%d")
all_trip$ToD <- format(all_trip$started_at, "%H:%M:%S")
```

-   The final steps involved removing any rows with negative ride
    lengths and empty rows in the start station name and end station
    name fields. To ensure that our original dataset is unchanged in
    case some chnages need to be reverted, we create a new dataset
    all_trip_v2.

```{r}
all_trip_v2<- all_trip %>% filter (!(ride_length < 0))

all_trip_v2<- all_trip_v2 %>% filter(!(is.na(start_station_name) | start_station_name == "")) %>% 
  filter(!(is.na(end_station_name) | end_station_name == ""))
```

## V Analysis Phase

The Analysis Phase is where we will aggregate our data, perform
calculations on it and identify trends and relationships that could be
useful towards accomplishing our business task.

To start with we perform some basic descriptive analysis on our data:

```{r}
mean(all_trip_v2$ride_length)
median(all_trip_v2$ride_length)
max(all_trip_v2$ride_length)
min(all_trip_v2$ride_length)
```

We notice that the average ride length is 1025.84 secs, the midpoint
number in the ascending array of ride lengths is 636 secs, maximum ride
length is 2057644 secs and minimum is 0 secs.

Next we look at the total number of each type of rider (casual and
annual member) and visualize it with the help of a bar chart:

```{r}
ggplot(all_trip_v2, aes(member_type, fill = member_type))+
  geom_bar()+geom_text(aes(label=..count..),stat = "count",vjust=1.5,color="black")
```

As illustrated above , there are 1758134 casual riders and 2611103
annual members present in the dataset.

Now we compare casual riders and annual members based on their ride
lengths:

```{r}
aggregate(all_trip_v2$ride_length ~ all_trip_v2$member_type, FUN = mean)
aggregate(all_trip_v2$ride_length ~ all_trip_v2$member_type, FUN = median)
aggregate(all_trip_v2$ride_length ~ all_trip_v2$member_type, FUN = max)
aggregate(all_trip_v2$ride_length ~ all_trip_v2$member_type, FUN = min)
```

As observed, casual riders have a longer ride length on average at
around **1440 secs** whereas annual members ride for an average of **747
secs**. Some other metrics noted were that casual riders have a median
ride length of **831 secs** compared to the **539 secs** of annual
members and the maximum ride length for them is **2057644 secs**
compared to **89594 secs** for annual members.

Having calculated the above information we can begin visualizing this
through a plot that illustrates the ride time per day of the week for
both types of riders:

```{r}
all_trip_v2$day_of_week <- ordered(all_trip_v2$day_of_week, 
          levels=c("Sunday", "Monday","Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

all_trip_v2 %>%
  group_by(member_type, day_of_week) %>%
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>%
  arrange(member_type, day_of_week) %>%
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_type)) +
  geom_col(position = "dodge")+
  labs(title = "Number of Rides per Day for both Member Types")
```

From this column chart we notice that casual riders ride most frequently
on the weekends and annual members are more likely to ride on the
weekdays. In fact on Saturdays, casual riders complete a greater number
of rides than annual members.

Having analyzed the rides per day for each member type we can now
proceed to take a look at the average ride length for each day of the
week:

```{r}
all_trip_v2 %>% group_by(member_type, day_of_week) %>% 
  summarise(avg_ride_length=mean(ride_length)) %>%
  arrange(member_type, day_of_week) %>%
  ggplot(aes(x=day_of_week, y=avg_ride_length, fill=member_type))+
  geom_col(position="dodge")+
  labs(title = "Average Ride length vs Day of Week for both Member Types")
```

On an average casual riders seem to ride bikes longer than annual
members.

To have a slightly deeper understanding of the riding habits of both
member types we take a look at two column charts, the first for number
of rides by member type per month and second for average ride length by
each member type per month:

Number of Rides per Month:

```{r}
all_trip_v2 %>% group_by(member_type, month) %>% 
  summarise(number_of_rides = n()) %>%
  arrange(member_type, month) %>%
  ggplot(aes(x=month, y=number_of_rides, fill=member_type))+
  geom_col(position="dodge")+
  labs(title = "Number of rides vs Month for both Member Types")
```

Average Ride Length per Month:

```{r}
all_trip_v2 %>% group_by(member_type, month) %>% 
  summarise(avg_ride_length=mean(ride_length)) %>%
  arrange(member_type, month) %>%
  ggplot(aes(x=month, y=avg_ride_length, fill=member_type))+
  geom_col(position="dodge")+
  labs(title = "Average Ride Length vs Month for both Member Types")
```

From our charts we can infer that peak ridership for casual riders is
between June to August and for annual members this trend extends up
until September. As we already know, the average ride length for casual
riders is greater overall than annual members but we also observe that
casual riders ride longer on average at the start of the year and then
drop off towards the end of the year. Annual members remain quite
consistent in terms of average ride length throughout the year.

Another interesting trend to take a look at would be the favorite
stations for each member type from where they begin their journeys. This
can be done as follows. First for Casual Riders:

```{r}
all_trip_v2 %>% filter(member_type == "casual") %>% 
  group_by(start_station_name) %>%
  summarise(n=n()) %>% arrange(desc(n)) %>% slice_max(n,n=10) %>%
ggplot(aes(x = start_station_name, y = n))+
  geom_col(position = "dodge", fill="orange")+
  theme(axis.text.x = element_text(angle = 45))+
  labs(title="Top 10 Start Stations for Casuals")
```

From the above graph, casual riders start most of their journeys from
**Streeter Dr & Grand Ave**

Then for Annual Members:

```{r}
all_trip_v2 %>% filter(member_type == "member") %>% 
  group_by(start_station_name) %>%
  summarise(n=n()) %>% arrange(desc(n)) %>% slice_max(n,n=10) %>%
  ggplot(aes(x = start_station_name, y = n))+
  geom_col(position = "dodge", fill="green")+
  theme(axis.text.x = element_text(angle = 45))+
  labs(title="Top 10 Start Stations for Members")
```

According to this analysis, annual members prefer to start from
**Kingsbury St & Kinzie St**

Having seen popular start stations among the different rider types,
analyzing the routes taken would also help us get a clearer picture of
how casual riders differ from annual members. For this first we create a
new column route by merging start station and end station names:

```{r}
all_trip_v2$bike_route = paste(all_trip_v2$start_station_name, all_trip_v2$end_station_name, 
                               sep = " to ")
```

Once done, we can continue to analyze favourite routes for our two rider
types:

Casual Riders:

```{r}
all_trip_v2 %>% filter(member_type == "casual") %>% 
  group_by(bike_route) %>%
  summarise(n=n()) %>% arrange(desc(n)) %>% slice_max(n,n=10) %>%
  ggplot(aes(x = bike_route, y = n))+
  geom_col(position = "dodge", fill="orange")+
  theme(axis.text.x = element_text(angle = 45))+
  labs(title="Most Popular Routes for Casuals")
```

The most popular route as observed above for casual riders is from
**Streeter Dr & Grand Ave** and back to the same station.

Annual Members:

```{r}
all_trip_v2 %>% filter(member_type == "member") %>% 
  group_by(bike_route) %>%
  summarise(n=n()) %>% arrange(desc(n)) %>% slice_max(n,n=10) %>%
  ggplot(aes(x = bike_route, y = n))+
  geom_col(position = "dodge", fill="green")+
  theme(axis.text.x = element_text(angle = 45))+
  labs(title="Most Popular Routes for Members")
```

The most popular route as observed above annual members is from **Ellis
Ave & 60th Street to University Ave & 57th Street**

For the final step of the analysis the bike type preferred by casual
riders and annual members is identified and plotted as follows:

```{r}
all_trip_v2 %>%
  group_by(member_type, rideable_type) %>%
  ggplot(aes(x = rideable_type, fill = member_type))+
  geom_bar(position = "dodge")+
  labs(title="Preferred bike types Casual vs Members")
```

This illustration clearly shows that annual members have a preference
for classic bikes whereas electric bikes are equally preferred by both
member types. There is no clear data on whether annual members use
docked bikes or not hence a comparison is not possible at this time.

## VI Conclusion

In this Case Study we began with a question that needed to be answered:
How do annual members and casual riders use Cyclistic bikes differently?
We then collected our data from a reputable data source and organized
and sorted it into appropriate folders in the Prepare phase. Then in the
Process phase we decided on R Studio as the tool of our choice and began
load our data, check for errors or inconsistencies and clean the data.
Finally in the Analysis phase we performed calculation on our data and
reached the following major conclusions that define the differences
between casual riders and Cyclistic annlual members:

-   Casual riders have a longer ride length on average at around **1440
    secs** whereas annual members ride for an average of **747 secs**.
    This could be because casual riders ride bikes for leisure or
    exercise whereas annual members might use the bikes for commuting
    purposes.

-   Casual riders ride most frequently on the **weekends** and annual
    members are more likely to ride on the **weekdays**. This again
    could be because of the contrasting purposes for which the members
    might use bikes.

-   Peak ridership for casual riders is between **June to August** and
    for annual members this trend extends up until **September**. The
    summer months are the most popular for bike riding as expected
    although annual members have a more even spread throughout the year
    due to the fact that they might be commuting by bikes.

-   Casual riders ride longer on average at the **start of the year**
    and then drop off towards the end of the year. Annual members remain
    quite **consistent** in terms of average ride length throughout the
    year. The beginning of the year often is when people are motivated
    more than usual to exercise more frequently and lead a healthier
    lifestyle which could be why casual riders ride longer at that time.
    The weather could also influence this trend.

-   Casual riders begin most of their rides from **Streeter Dr & Grand
    Ave** and their favorite route in turn is from Streeter Dr & Grand
    Ave back to the same station. Annual members start most frequently
    from **Kingsbury St & Kinzie St** but their most used route is from
    **Ellis Ave & 60th Street to University Ave & 57th Street**.

-   Lastly, Annual members have a preference for **classic bikes**
    whereas **electric bikes** are almost equally preferred by both
    member types. Here an assumption could be made that either annual
    members belong to an older age group that prefer to not only commute
    by bikes but also prefer classic bikes over electric or that the
    stations they start from have more number of classic bikes
    available.

